\name{}: Combining Bottom-up and top-down approaches through graph learning over interaction networks for drug-target-interaction prediction

### 2021-04-13 paper/main.tex
Update citations and fix some TODOs

Both bottom-up and top-down approaches to drug--target interaction
prediction have some advantages and limitations. Generally, bottom-up
methods face the challenge to predict whether a structure binds to a
protein given their molecular properties; whether two entities
interact depends not only on the molecular structure of the entities
(where binding sites and molecular forces need to be determined for
accurate prediction) but also properties such as where a protein is
expressed or in which celltypes and anatomical structures. Top-down
methods use information for DTI prediction such as side-effect
similarity \cite{Campillos2008} that is largely complementary to
knowledge gained from molecular properties. While methods that rely on
molecular information are directly predicting whether two molecular
could interact, top-down methods rely on more indirect means and 
infer a DTI from observable effects resulting from the interaction.

Both approaches may be used in conjunction with network inference
\cite{}. Biological networks used for DTI prediction include
protein--protein interaction networks \cite{} and networks including
several other types of biological relations, including similarity
between represented entities \cite{PREDICT}. Network-based DTI
prediction methods use the guilt-by-association principle
\cite{https://journals.plos.org/ploscompbiol/article?id=10.1371/journal.pcbi.1002444}
and assume that a protein is a likely target for a drug if many of the
protein's neighbors in the interaction network are targets of the drug
\cite{https://academic.oup.com/bioinformatics/article/32/7/1057/1743938,
  others}. Network-based methods have been applied successfully
to DTI prediction. However, if DTIs are taken as direct physical
interactions between a drug and protein, it remains an unresolved
question whether the network-based guilt-by-association hypothesis is
true, or whether an interaction of a drug and protein dys-regulates
several of the protein's interaction partners, and therefore resulting
in downstream effects that are not direct interactions but only
downstream consequences of an interaction.

Progress in machine learning using graph neural networks can allow us
to test this hypothesis and combine both bottom-up and top-down
features with a network in a single machine learning model. 
In particular, Graph Convolutional Networks \cite{GCNConv} and
their variants using different types of kernels \cite{ChebConv,
  ARMAConv}, including attention mechanisms \cite{GATConv}, and
different forms of exploring node neighborhoods \cite{APPNPConv,
  SAGEConv} can combine different types of features and graph-based
information. % While based on diverse systems,
% they can be relevant for testing distinct hypothesis for given
% graphs. While convolutional filters are suitable for finding patterns
% among the given graph, attention mechanisms are more relevant for
% discovery of important regions within. Lately, graph learning
% approaches found application for computing compound representations
% for DTI prediction.
They have previously been applied for a number of tasks, including
prediction of protein functions \cite{Zitnik2017}, \todo{add 2 more,
  and then one on DTI prediction}.

@@

We developed \name{} as a method for predicting DTIs. We use an
ontology-based machine learning method \cite{} to encode phenotypic
consequences of DTIs and deep learning methods to encode molecular
features. We combine both using a protein interaction network which we
exploit using a graph neural network. We use this model to test
whether molecular or phenotype features benefit from the network
information and find that only phenotype features localize on the
graph whereas molecular features do not. We further evaluate and
compare \name{} against several DTI prediction methods and
demonstrate a substantial improvement of \name{} over the state of the
art in predicting drugs that target a protein. We also identify and
characterize several biases in both training and evaluating DTI
prediction methods, and make recommendations on how to avoid
them. \name is available as Free Software at \url{}\todo{add}.

@@

We utilize the PhenomeNET ontology \citep{PhenomeNET2011}, an ontology
integrating ontologies such as the Human Phenotype Ontology
\citep{HPO2018}, Gene Ontology \cite{GOoriginal2000, GOrecent2020},
Mammalian Phenotype Ontology \citep{MP2009} and several others.  We
obtained side effects and their links to drugs from SIDER
\citep{SIDER}; SIDER contains side effects encoded using identifiers
from the MedDRA database \citep{MedDRA}. We mapped side effects to the
PhenomeNET ontology using the \textit{Phenomebrowser.net}, which
provides a SPARQL \cite{} query endpoint for the mentioned resources.

@@

For comparative evaluation, we use the gold standard dataset
introduced by \citet{Yaminishi2008} consisting of 1,923 interactions
between 708 drugs and 1,512 proteins, and the BioSnap dataset
\cite{BioSnap2018} which consists of 5,017 drug nodes, 2,324 gene
nodes and 15,138 edges.

@@

We trained, validated and finally tested all considered models on the
STITCH dataset using a 5-fold cross-validation over a protein split;
we then selected the best-performing models (with respect to
$MicroAUC_p$, see Section \ref{sec:evaluation}) and evaluate\todo{Only
evaluate or retrain (from scratch)? Should say here.} them in a
5-fold protein-split cross-validation on the Yamanishi benchmark
dataset to avoid validation overfitting and yield more realistic
testing results. To evaluate the influence of the different features
separately, and to determine whether they ``localize'' on the PPI
graph (and therefore can be exploited successfully by the graph neural
networks), we train and evaluate models with different types of
features, and with and without inclusion of the PPI graph,
separately. We comparing the molecular (MolPred) and phenotype-based
(OntoPred) prediction model, and a combination of both where we
concatenate both types of features.  Table \ref{tab:Results} shows the
results of these experiments.

@@

The improvements when including the graph are only provided by the
\textit{GENConv} graph convolution scheme which includes the
{\em ResGraphConv} blocks; \textit{GCNConv} and other graph
convolutional methods fail to achieve any gain in comparison to the
plain \textit{OntoPred} performance even when combined with the
residual blocks. The discrepancy between {\em GENConv} and other graph
convolutional methods may be the result of numerical instability and
fading signals \cite{}\todo{Add citation to original paper}.

@@

We apply this naive predictor on both the STITCH and Yamanishi
datasets, using the full datasets as well as a 5-fold cross-validation
over drugs and over drug--protein pairs to compare the prediction
results directly to \name{}.  For each fold, we gradually increase $k$
to determine the best performance for each fold. Using the full
dataset, drug--target split, and drug split, we obtain the following
MacroAUC results: for the STITCH database, we obtain a performance of
$0.76$ on the whole dataset, $0.70$ for the drug--target pairs and
$0.73$ in case of the drug splitting scheme; on the Yamanishi dataset,
we obtain MacroAUC scores of $0.88$, $0.84$ and $0.85$, for the total
dataset, drug--target pair and drug split, respectively\todo{In the
  Table is says 0.85 on Yaminishi for DP pair split, here it says 0.84
  -- which is correct? Need fix}.  The na\"ive
predictor shows higher performance on the Yamanishi dataset than on
STITCH, and a substantial gain in comparison to an expected random
predictor on both datasets. In the following, we utilize this nai\"ve
predictor as baseline to compare its performance to state of the art
models and \name{}.

@@

    &Splitting scheme&Macro AUC&Macro AUC&Micro $AUC_p$\\
    \hline
    Naive predictor&DP pairs&$0.85$& --&--\\
    DTINet&DP pairs&$0.91$&$0.74$&$0.67$ \\
    DTIGEMS+&DP pairs&$\mathbf{0.93}$& $0.72$& $0.68$ \\
    DTI-CDF&Proteins&$0.85$&$\mathbf{0.85}$&$0.79$\\
    \name&Proteins&$0.84$&$0.84$&$\mathbf{0.84}$\\
    \hline
  \end{tabular}
  \caption{\label{tab:comparison1} Comparison of \name{} with
    state of the art drug--target interaction prediction methods on
    the Yamanishi dataset; we evaluated the original and the
    protein-based split in a cross-validation}.
\end{table}

>>>>

Both bottom-up and top-down approaches to drug--target interaction
prediction have some advantages and limitations. Generally, bottom-up
methods face the challenge to predict whether a structure binds to a
protein given their molecular properties; whether two entities
interact depends not only on the molecular structure of the entities
(where binding sites and molecular forces need to be determined for
accurate prediction) but also properties such as where a protein is
expressed or in which celltypes and anatomical structures. Top-down
methods use information for DTI prediction such as side-effect
similarity \cite{Campillos2008} that is largely complementary to
knowledge gained from molecular properties. While methods tha3t rely on
molecular information are directly predicting whether two molecular
could interact, top-down methods rely on more indirect means and 
infer a DTI from observable effects resulting from the interaction.

Both approaches may be used in conjunction with network inference
\cite{}. Biological networks used for DTI prediction include
protein--protein interaction networks \cite{} and networks including
several other types of biological relations, including similarity
between represented entities \cite{Gottlieb2011}. Network-based DTI
prediction methods use the guilt-by-association principle
\cite{Oliver2000}
and assume that a protein is a likely target for a drug if many of the
protein's neighbors in the interaction network are targets of the drug
\cite{Gillis2012}. Network-based methods have been applied successfully
to DTI prediction. However, if DTIs are taken as direct physical
interactions between a drug and protein, it remains an unresolved
question whether the network-based guilt-by-association hypothesis is
true, or whether an interaction of a drug and protein dys-regulates
several of the protein's interaction partners, and therefore resulting
in downstream effects that are not direct interactions but only
downstream consequences of an interaction.

Progress in machine learning using graph neural networks can allow us
to test this hypothesis and combine both bottom-up and top-down
features with a network in a single machine learning model. 
In particular, Graph Convolutional Networks \cite{GCNConv} and
their variants using different types of kernels \cite{ChebConv,
  ARMAConv}, including attention mechanisms \cite{GATConv}, and
different forms of exploring node neighborhoods \cite{APPNPConv,
  SAGEConv} can combine different types of features and graph-based
information. % While based on diverse systems,
% they can be relevant for testing distinct hypothesis for given
% graphs. While convolutional filters are suitable for finding patterns
% among the given graph, attention mechanisms are more relevant for
% discovery of important regions within. Lately, graph learning
% approaches found application for computing compound representations
% for DTI prediction.
They have previously been applied for a number of tasks, including
prediction of protein functions \cite{Zitnik2017}, cancer drug response \cite{Liu2020} and drug--target affinity prediction \cite{GraphDTA2020}.

@@

We developed \name{} as a method for predicting DTIs. We use an
ontology-based machine learning method \cite{DL2vec2020} to encode phenotypic
consequences of DTIs and deep learning methods to encode molecular
features. We combine both using a protein interaction network which we
exploit using a graph neural network. We use this model to test
whether molecular or phenotype features benefit from the network
information and find that only phenotype features localize on the
graph whereas molecular features do not. We further evaluate and
compare \name{} against several DTI prediction methods and
demonstrate a substantial improvement of \name{} over the state of the
art in predicting drugs that target a protein. We also identify and
characterize several biases in both training and evaluating DTI
prediction methods, and make recommendations on how to avoid
them. \name{} is available as Free Software at \url{}\todo{add}.

@@

We utilize the PhenomeNET ontology \citep{PhenomeNET2011}, an ontology
integrating ontologies such as the Human Phenotype Ontology
\citep{HPO2018}, Gene Ontology \cite{GOoriginal2000, GOrecent2020},
Mammalian Phenotype Ontology \citep{MP2009} and several others.  We
obtained side effects and their links to drugs from SIDER
\citep{SIDER}; SIDER contains side effects encoded using identifiers
from the MedDRA database \citep{MedDRA}. We mapped side effects to the
PhenomeNET ontology using the \textit{Phenomebrowser.net}, which
provides a SPARQL query endpoint for the mentioned resources.

@@

For comparative evaluation, we use the gold standard dataset
introduced by \citet{Yamanishi2008} consisting of 1,923 interactions
between 708 drugs and 1,512 proteins, and the BioSnap dataset
\cite{BioSnap2018} which consists of 5,017 drug nodes, 2,324 gene
nodes and 15,138 edges.

@@

We trained, validated and finally tested all considered models on the
STITCH dataset using a 5-fold cross-validation over a protein split;
we then selected the best-performing models (with respect to
$MicroAUC_p$, see Section \ref{sec:evaluation}), and retrain them from scratch in a
5-fold protein-split cross-validation on the Yamanishi benchmark
dataset to avoid validation overfitting and yield more realistic
testing results. To evaluate the influence of the different features
separately, and to determine whether they ``localize'' on the PPI
graph (and therefore can be exploited successfully by the graph neural
networks), we train and evaluate models with different types of
features, and with and without inclusion of the PPI graph,
separately. We comparing the molecular (MolPred) and phenotype-based
(OntoPred) prediction model, and a combination of both where we
concatenate both types of features.  Table \ref{tab:Results} shows the
results of these experiments.

@@

The improvements when including the graph are only provided by the
\textit{GENConv} graph convolution scheme which includes the
{\em ResGraphConv} blocks; \textit{GCNConv} and other graph
convolutional methods fail to achieve any gain in comparison to the
plain \textit{OntoPred} performance even when combined with the
residual blocks. The discrepancy between {\em GENConv} and other graph
convolutional methods may be the result of numerical instability and
fading signals \cite{DeeperGCN2020}.

@@

We apply this naive predictor on both the STITCH and Yamanishi
datasets, using the full datasets as well as a 5-fold cross-validation
over drugs and over drug--protein pairs to compare the prediction
results directly to \name{}.  For each fold, we gradually increase $k$
to determine the best performance for each fold. Using the full
dataset, drug--target split, and drug split, we obtain the following
MacroAUC results: for the STITCH database, we obtain a performance of
$0.76$ on the whole dataset, $0.70$ for the drug--target pairs and
$0.73$ in case of the drug splitting scheme; on the Yamanishi dataset,
we obtain MacroAUC scores of $0.88$, $0.84$ and $0.85$, for the total
dataset, drug--target pair and drug split, respectively.  The na\"ive
predictor shows higher performance on the Yamanishi dataset than on
STITCH, and a substantial gain in comparison to an expected random
predictor on both datasets. In the following, we utilize this nai\"ve
predictor as baseline to compare its performance to state of the art
models and \name{}.

@@

    &Splitting scheme&Macro AUC&Macro AUC&Micro $AUC_p$\\
    \hline
    Naive predictor&Drugs&$0.85$& --&--\\
    DTINet&DP pairs&$0.91$&$0.74$&$0.67$ \\
    DTIGEMS+&DP pairs&$\mathbf{0.93}$& $0.72$& $0.68$ \\
    DTI-CDF&Proteins&$0.85$&$\mathbf{0.85}$&$0.79$\\
    \name&Proteins&$0.84$&$0.84$&$\mathbf{0.84}$\\
    \hline
  \end{tabular}
  \caption{\label{tab:comparison1} Comparison of \name{} with
    state of the art drug--target interaction prediction methods on
    the Yamanishi dataset; we evaluated the original and the
    protein-based split in a cross-validation}.
\end{table}
